---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dpi = 144
)
```

# ggcorrheatmap

<!-- badges: start -->
[![R-CMD-check](https://github.com/leod123/ggcorrheatmap/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/leod123/ggcorrheatmap/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

ggcorrheatmap is a convenient package for generating correlation heatmaps made with ggplot2, with support for triangular layouts, clustering and annotation. As the output is a ggplot2 object you can further customise the appearance using familiar ggplot2 functions. ggcorrheatmap can also build general heatmaps.

## Installation

You can install the development version of ggcorrheatmap from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("leod123/ggcorrheatmap")
```

## Example

Below is an example of how to generate a correlation heatmap with clustered rows and columns and row annotation, using a triangular layout that excludes redundant cells.

```{r example1, out.width='70%', fig.align='center'}
library(ggcorrheatmap)

set.seed(123)
# Make a correlation heatmap with a triangular layout, annotations and clustering
row_annot <- data.frame(.names = colnames(mtcars),
                        annot1 = sample(letters[1:3], ncol(mtcars), TRUE),
                        annot2 = rnorm(ncol(mtcars)))
ggcorrhm(mtcars, layout = "bottomright",
         cluster_rows = TRUE, cluster_cols = TRUE,
         show_dend_rows = FALSE, annot_rows_df = row_annot)

```

Or a mixed layout that displays different things in the different triangles.

```{r example2, out.width='70%', fig.align='center'}
# With correlation values and p-values
ggcorrhm(mtcars, layout = c("topright", "bottomleft"),
         cell_labels = c(FALSE, TRUE), p_values = c(FALSE, TRUE))
```


It is also possible to make a normal heatmap, for a more flexible output.

```{r example3}
gghm(scale(mtcars), cluster_rows = TRUE, cluster_cols = TRUE)
```

