---
title: "Scales and legends"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, results='hide', message=FALSE, warning=FALSE}
library(ggcorrheatmap)
library(ggplot2)
library(patchwork)  # Arrange plots and legends
library(cowplot)    # For extracting and arranging legends
```

# Scales

As explained in some of the other articles, the colour scale argument `col_scale` takes NULL (for default scales), a character (for Brewer and Viridis scales) and `ggplot2` scale objects. When passing scale objects, make sure the correct aesthetic is used (fill or colour/color depending on plotting mode).

```{r, fig.alt='Three heatmaps of the volcano data. The first one uses the default dark blue to light blue gradient for the cells, the second one uses the Viridis magma palette, and the third a custom scale going from light blue to white to pink.'}
plt1 <- gghm(volcano, col_name = "Default",
             border_col = NA, show_names_x = FALSE, show_names_y = FALSE)

plt2 <- gghm(volcano,
             # Brewer Spectral palette
             col_scale = "A", col_name = "Character",
             border_col = NA, show_names_x = FALSE, show_names_y = FALSE)

plt3 <- gghm(volcano,
             # Custom scale
             col_scale = scale_fill_gradient2(
               high = "pink", mid = "white", low = "lightblue",
               midpoint = 140, name = "Custom"
             ),
             border_col = NA, show_names_x = FALSE, show_names_y = FALSE)

plt1 + plt2 + plt3 &
  theme(legend.position = "bottom",
        legend.title = element_text(vjust = 0.8))
```

This is also true for the annotation scale arguments `annot_rows_col` and `annot_cols_col`.

```{r, fig.height=5.2, fig.width=5, fig.alt='A heatmap of the first ten rows and five columns of mtcars. Three row annotations can be seen on the right side, the first one using the Viridis scale, the second one the Brewer Spectral, and the third one a custom scale going from light blue to white to pink.'}
row_annot <- data.frame(.names = rownames(mtcars)[1:10],
                        Default = 1:10,
                        Character = 1:10,
                        Custom = 1:10)

row_annot_fill <- list(Default = NULL,
                       Character = "Spectral",
                       Custom = scale_fill_gradient2(
                         high = "pink", mid = "white", low = "lightblue",
                         midpoint = 6
                       ))

gghm(mtcars[1:10, 1:5],
     annot_rows_df = row_annot,
     annot_rows_col = row_annot_fill,
     # Make annotation larger and hide heatmap legend
     annot_size = 1,
     legend_order = c(NA, 1:3)) +
  theme(plot.margin = margin(0, 0, 30, 0))
```

When `col_scale` is a character, "rev_" or "_rev" can be added to the beginning or end, respectively, to reverse the scale. This also works for annotation scales.

```{r, fig.height=6, fig.width=5, fig.alt='A heatmap of the volcano data with cells coloured using the Viridis magma palette but reversed so that low values are light and high are dark. Two row annotations are seen to the right, both have the same values but one is using the default magma palette and the other the reversed one.'}

row_annot <- data.frame("Default" = 1:nrow(volcano),
                        "Reversed" = 1:nrow(volcano))
row_annot_colr <- list("Default" = "A", "Reversed" = "A_rev")

gghm(volcano, show_names_x = FALSE, show_names_y = FALSE, border_col = NA,
     col_scale = "rev_A", col_name = "Reversed",
     annot_rows_df = row_annot,
     annot_rows_col = row_annot_colr,
     # Make the annotation larger for visibility
     annot_size = 3, annot_dist = 1.5) +
  # Change the margins a bit too
  theme(plot.margin = margin(0, 0, 40, 0))
```


The default scale for the main heatmap is different for `ggcorrhm()` as it assumes that correlations are plotted. There are also more arguments to modify the scale, such as the colours or binning.

When the plot is using a [mixed layout](mixed.html), `col_scale` can also take a vector or list to apply different scales to the different triangles.

```{r, fig.alt='A heatmap of a 10x10 matrix with values arranged so that there is a gradient across the heatmap with low values in the bottom left and high values in the top right. The top left triangle uses the Viridis magma scale, while the bottom right triangle is using the Brewer RdYlGn palette.'}
plt_dat <- sapply(1:10, function(x) seq(x, x + 9))

gghm(plt_dat, layout = c("tl", "br"),
     mode = c("hm", "hm"),
     # For Viridis options, both the short and long versions work
     col_scale = c("magma", "RdYlGn"))
```

`col_scale` can still be one value to apply the same scale to both triangles. This does not work for scale objects, as the two triangles may use different aesthetics (fill or colour). This means that there will be two colour legends for the main heatmap if the two triangles use different aesthetics, even if they use the same colours. `ggcorrhm()` hides one of the legends by default and the [legend order section](legends.html#legend-order) goes into more detail on how this can be done.

```{r, fig.alt='A heatmap of the same gradient data as the previous figure. The bottom right triangle now has coloured numbers (cell values) in the cells, using the same colour palette (Viridis magma) as the top left triangle. Two legends showing the same values and scale are on the right.'}
gghm(plt_dat, layout = c("tl", "br"),
     mode = c("hm", "text"), col_scale = "A")
```

As also explained in the [mixed layout](mixed.html#mixed-scales) article, the scale-modifying arguments `bins`, `na_col`, `limits`, and in `ggcorrhm()` also `high`, `mid`, `low`, `midpoint`, and `size_range` can be applied triangle-wise. If only one scale is provided but a scale-modifying argument is applied differently per triangle, two scales are created anyway.

```{r, fig.alt='A 10x10 heatmap showing a gradient of values going from low values (starting at 1) in the top left to high values (up to 19) in the bottom right. The Viridis mako colour scale is used, shifting from black (low values) through blue and teal and ending at a pale green. The top right triangle of the heatmap is divided in to four bins while the bottom left triangle has eight.'}
# Only one scale supplied, but it is split into two by
# specifying two different binnings
gghm(plt_dat, layout = c("tr", "bl"), mode = c("hm", "hm"),
     col_scale = "G", bins = c(4L, 8L))
```

When using scale objects, arguments that change scales and legends such as `col_name` or `legend_order` and, for `ggcorrhm()`, `high`, `mid`, `low` etc will be ignored to instead give the user full control. This allows for more flexibility, such as changing the breaks, bin sizes, or changing how out of bounds values are treated.

The default way `ggplot2` deals with values outside of the scale range (out of bound value) is to replace them with NAs. The `oob` argument of the scale functions can change the behaviour, e.g. so that OOB values use the extreme colours of the scale.

```{r, fig.height=5, fig.width=10, fig.alt='Two figures of the gradient plot from before, now with low numbers in the top left and high numbers in the bottom right. The values are written in the cells with black text, showing that the gradient goes from 1 to 19. The first plot is simply coloured using the Brewer RdYlGn palette. The second plot uses the same palette, but the limits have been set to 7 and 13, narrowing the colour gradient that was visible in the first plot. Values outside the range are beige in the top right triangle, and green or red in the bottom left (corresponding to the lowest and highest ends of the scale).'}
# Normal limits
plt1 <- gghm(plt_dat, col_scale = "RdYlGn_rev",
             cell_labels = TRUE)

# Narrower limits with handling of oob values
plt2 <- gghm(plt_dat, layout = c("tr", "bl"),
     mode = c("hm", "hm"),
     col_scale = list(
       # scale_fill_distiller to use Brewer palettes with continuous data
       # The default direction in distiller is -1, opposite of gghm
       scale_fill_distiller(
         palette = "RdYlGn",
         # Values outside the limits are replaced with NAs
         limits = c(7, 13),
         # Can then change the colour using the na.value argument
         # (corresponding to na_col in gghm)
         na.value = "beige"
       ),
       
       scale_fill_distiller(
         palette = "RdYlGn",
         limits = c(7, 13),
         # Change oob strategy 'oob' argument
         # scales::squish assigns the extreme colours to values outside the range
         oob = scales::squish
       )
     ), cell_labels = TRUE)

plt1 + plt2
```

When binning scales with the `bins` argument, a number of the 'double' class will prioritise breaks at nice numbers while an integer will prioritise the number of bins. As a side note, out-of-bound values with binned scales also default to using the 'squish' strategy shown above.

```{r, fig.height=5, fig.width=10, fig.alt='Two 100x100 heatmaps with values from -1 to 1 in a gradient going from the bottom left to the top right. The heatmaps are using the Brewer Spectral colour scale in the top left triangle and the Viridis turbo scale in the bottom right. The first heatmap simply shows the gradient. The second heatmap has binned scales, with 11 bins in the top left triangle (with breaks at numbers with many decimals and limits at -0.5 and 0.5) and 10 bins in the bottom right triangle (evenly spaced at intervals of 0.2 going from -1 to 1).'}

# Make plot data between -1 and 1 (like correlation data)
plt_dat_cor <- sapply(seq(-1, 0, length.out = 100), function(x) {
  seq(x, x + 1, length.out = 100)
})

# Simple mixed layout with mixed scales
plt1 <- ggcorrhm(plt_dat_cor, cor_in = TRUE, show_names_diag = FALSE,
                 border_col = 0, layout = c("tl", "br"), mode = c("hm", "hm"),
                 col_scale = c("Spectral", "H"))

# With many scale arguments applied
plt2 <- ggcorrhm(plt_dat_cor, cor_in = TRUE, show_names_diag = FALSE,
                 border_col = 0, layout = c("tl", "br"), mode = c("hm", "hm"),
                 col_scale = c("Spectral", "H"),
                 # Provide a list as a mix of integer and float would be coerced if atomic vector
                 # Integer vs double, results in 11 and 10 bins
                 bins = list(11L, 11),
                 # NULL to use default limits (-1 to 1 in ggcorrhm)
                 limits = list(c(-.5, .5), NULL))

plt1 + plt2

```

By using scale objects, breaks in both continuous and binned scales can be put wherever desired.

```{r, fig.height=5, fig.width=10, fig.alt='The same kind of gradient heatmaps as shown in the previous plots, but the values now range from -1 (top left) to 1 (bottom right). A diverging colour scale going from blue to white to red is used for the cell colours. The first plot simply shows off the gradient, while the second uses two binned scales ranging between -0.7 to 0.7. The scale in the top right triangle has five bins with breaks at 0.42, 0.14, -0.42, and -0.14, while the bottom left triangle has four bins with breaks at 0.4, 0, and -0.4.'}

plt_dat_cor <- sapply(seq(-1, 0, length.out = 10), function(x) {
  seq(x, x + 1, length.out = 10)
})

plt1 <- ggcorrhm(plt_dat_cor, cor_in = TRUE)

plt2 <- ggcorrhm(plt_dat_cor, cor_in = TRUE,
                 layout = c("tr", "bl"),
                 mode = c("hm", "hm"),
                 col_scale = list(
                   scale_fill_gradient2(
                     # Mimic the default correlation scale
                     high = "sienna2", mid = "white", low = "skyblue2",
                     # If not binned this only changes the legend but not the colours
                     breaks = seq(-1, 1, length.out = 11)
                   ),
                   # Binned scales can be made with e.g.
                   # scale_fill_fermenter, scale_fill_viridis_b, scale_fill_steps etc
                   scale_fill_steps2(
                     high = "sienna2", mid = "white", low = "skyblue2",
                     # Specify exact break points for bins
                     breaks = c(-1, -0.75, -0.25, 0, 0.25, 0.75, 1)
                   )
                 ))

plt1 + plt2
```

Size scales can be customised as well. `ggcorrhm()` uses an absolute value transform for the default size scale and hides the legend as it becomes inaccurate (as can be seen in the plot below). `scale_size_area` can be passed to `size_scale` to still scale with the absolute value and keep a meaningful legend, at the cost of not being able to set the lower limit for the sizes.

```{r, fig.alt='mtcars correlation heatmap with diamonds instead of cells. In the top right triangle the diamonds are smaller than in the bottom left. The sizes of all diamonds scale with the absolute value of the correlations, but the legend for the top right diamonds goes from 0 to 1 while the legend for the other triangle properly shows the negative values too.'}
ggcorrhm(mtcars, layout = c("tr", "bl"), mode = c(23, 23),
         size_scale = list(
           # Can make a binned size scale
           scale_size_binned(n.breaks = 6, range = c(2, 6),
                             # Absolute value transform but legend loses meaning
                             transform = scales::trans_new("abs", abs, abs),
                             name = "scale_size_binned\n(absolute value)",
                             guide = guide_bins(order = 1)),
           # scale_size_area, scale_size_binned_area also scale with absolute value
           # but fewer options to control sizes
           scale_size_area(max_size = 10, name = "scale_size_area")
         ), border_lwd = 1, legend_order = NA)
```


# Legend order

Legends have a set default order:

1. Main heatmap colours.
2. Main heatmap sizes.
3. Row annotations in order.
4. Column annotations in order.

```{r, fig.alt='A heatmap of the first 10 rows of mtcars (scaled), with coloured diamonds instead of cells. The heatmap colours range from dark blue to light blue. There is one row annotation (using the viridis D option) and one column annotation (using the viridis A option), named 3 and 4, respectively. Four legends are on the right side of the plot for the main heatmap colours, main heatmap sizes, row annotations, and column annotations. The legends are named 1, 2, 3, and 4.'}
row_annot <- data.frame(.names = rownames(mtcars)[1:10], 1:10)
colnames(row_annot)[2] <- "3"

col_annot <- data.frame(.names = colnames(mtcars), 11:1)
colnames(col_annot)[2] <- "4"

gghm(scale(mtcars)[1:10, ], mode = 23,
     col_name = "1", size_name = "2",
     annot_rows_df = row_annot,
     annot_cols_df = col_annot) +
  # Make the legends horizontal to fit better
  theme(legend.direction = "horizontal")
```

If the layout is mixed, the legends of the first triangle appear before those of the second triangle.

```{r, fig.alt='mtcars correlation heatmap with cells in the bottom left triangle and squares with sizes varying with the value in the top right. The deafult ggplot2 colour scale (dark blue to light blue) is used for both triangles. Three legends are on the right, the first one a colour scale, the second one a size scale, and the third one another colour scale.'}
gghm(cor(mtcars), layout = c("tr", "bl"), mode = c("15", "hm"),
     col_scale = c("H", "A"), col_name = c("Triangle 1", "Triangle 2"),
     size_name = "Triangle 1") +
  theme(legend.direction = "horizontal")
```

The order can be changed by using the `legend_order` argument. This argument takes a numeric vector, the first element specifying the order for the first legend (of the default order), the second element for the second legend and so on. If a value is `NA`, that legend is hidden.

```{r, fig.width=8, fig.height=10, fig.alt='Two mtcars correlation heatmaps with circles instead of cells. Both have two row and two column annotations each. The first heatmap has seven legends on the right (colour, size, colour, row annotation colours 1-2, column annotation colours 1-2), named 1 to 7. The second heatmap is the same as the first but with the legends ordered 4, 2, 7, 1, 5. The 3 and 6 legends are missing.'}
set.seed(123)
annot_rows <- data.frame(.names = colnames(mtcars), 1:11, 11:1)
colnames(annot_rows)[2:3] <- c("4", "5")
annot_cols <- data.frame(.names = colnames(mtcars),
                         sample(letters[1:3], 11, TRUE),
                         sample(LETTERS[1:2], 11, TRUE))
colnames(annot_cols)[2:3] <- c("6", "7")

plt1 <- gghm(cor(mtcars), layout = c("tl", "br"), mode = c(19, 21),
             col_name = c("1", "3"), size_name = c("2"),
             annot_rows_df = annot_rows,
             annot_cols_df = annot_cols)

plt2 <- gghm(cor(mtcars), layout = c("tl", "br"), mode = c(19, 21),
             col_name = c("1", "3"), size_name = c("2"),
             annot_rows_df = annot_rows,
             annot_cols_df = annot_cols,
             legend_order = c(4, 2, NA, 1, 5, NA, 3))

plt1 / plt2 &
  theme(legend.direction = "horizontal")
```

A single NA value hides all legends.

```{r, fig.alt='The mtcars correlation heatmap with circles as above, but with no legends.'}
gghm(cor(mtcars), layout = c("tl", "br"), mode = c(19, 21),
     col_name = c("1", "3"), size_name = c("2"),
     annot_rows_df = annot_rows,
     annot_cols_df = annot_cols,
     legend_order = NA)
```

Scales can be specified as `ggplot2` scale objects. In this case the user has complete control over the scale and it is unaffected by arguments like `legend_order` or `col_name`.

```{r, fig.alt='mtcars correlation heatmap with circles like above but with some colours altered. The top left triangle uses the dark to light blue colour scale, while the bottom right triangle uses the Viridis mako scale (black to teal or emerald-like green). The second row annotation uses the Brewer Purples scale (white to purple). The second column annotation has red, white, and blue. Only the bottom right triangle and second row annotation have legends to the right.'}
gghm(cor(mtcars), layout = c("tl", "br"), mode = c(19, 21),
     col_name = c("1", "3"), size_name = "2",
     col_scale = list(
       # Keep default scale for first triangle
       NULL,
       # Pass a scale object for second triangle
       scale_fill_viridis_c(name = "scale object 1", option = "G",
                            guide = guide_colourbar(order = 1))
     ),
     annot_rows_df = annot_rows,
     annot_cols_df = annot_cols,
     annot_rows_col = list(
       # Scale object for second row annotation
       "5" = scale_fill_distiller(name = "scale object 2",
                                  palette = "Purples")
     ),
     annot_cols_col = list(
       # Scale object for first column annotation
       "6" = scale_fill_brewer(name = "scale object 3",
                               palette = "RdBu",
                               # Hide the legend
                               guide = "none")
     ),
     # (Try to) Remove all legends
     legend_order = NA) +
  theme(legend.direction = "horizontal")
```

# Legends in correlation heatmaps

`ggcorrhm()` tries to hide some legends by default when using mixed layouts to reduce clutter.

For example, when both triangles use the same colour scale, one of the scales is hidden. This only happens when the default scale is used (customised with `high`, `mid`, etc is ok) or when `col_scale` is a character. If `col_scale` contains `ggplot2` scale objects both legends will be shown (unless specified not to in the scale object).

Default size legends are also hidden as they use an absolute value transformation and show different information from the colour scale.

```{r, fig.alt='mtcars correlation heatmap with cells in the top left and diamonds in the bottom right. The heatmap uses the Brewer RdYlBu scale that ranges from blue to yellow to red. There is only one legend on the right, showing the colour scale.'}
# If both triangles use the same colour scale
# (when col_scale is NULL or a one character value)
ggcorrhm(mtcars, layout = c("tl", "br"),
         mode = c("hm", "18"), col_scale = "RdYlBu")
```

The legends can be included again by using the `legend_order` argument or using scale objects for the scales.

```{r, fig.alt='mtcars correlation heatmap with cells and diamonds and the RdYlBu scale like above. There are three legends one the right, the first two for the colour scale and the last one for the sizes (showing the absolute values).'}
ggcorrhm(mtcars, layout = c("tl", "br"),
         mode = c("hm", "18"), col_scale = "RdYlBu",
         # Elements beyond the number of legends are ignored
         # On a related note, if legend_order is shorter than
         # the number of legends, legends not in the vector are hidden
         legend_order = 1:10)
```

# Arranging legends

When there are many legends, some of them may be cut off and it can be tricky to make them fit. One way is to change the positions, sizes or orientations of legends as shown in some of the plots above. Another way is to change the size of the figure until the legends fit. A more advanced way is to use packages like `cowplot` and `patchwork` to stitch together a legend-less version of the plot with the legends, which also makes it possible to arrange the legends in grids.

```{r, fig.alt='mtcars correlation heatmap with clustered rows and columns, as well as two row annotations and one column annotation. Four legends can be seen to the right, but they do not all fit and the first and last one cannot be seen fully.'}

# Plot with legends
p <- ggcorrhm(mtcars, cluster_rows = TRUE, cluster_cols = TRUE,
              annot_rows_df = data.frame(.names = colnames(mtcars),
                                         a = 1:11, b = 11:1),
              annot_cols_df = data.frame(.names = colnames(mtcars),
                                         c = 1:11))

# Plot without legends
pl <- ggcorrhm(mtcars, cluster_rows = TRUE, cluster_cols = TRUE,
               annot_rows_df = data.frame(.names = colnames(mtcars),
                                          a = 1:11, b = 11:1),
               annot_cols_df = data.frame(.names = colnames(mtcars),
                                          c = 1:11),
               legend_order = NA)

# Show plot with default legends
p
```

```{r, fig.alt='The same heatmap with annotations as the previous figure, but the legends are now all visible, arranged in a 2x2 grid to the right of the plot.'}
# Extract the four legends
lgd <- get_legend(p)[["grobs"]][1:4]

# Arrange in a grid (cowplot)
lgd_grid <- plot_grid(plotlist = lgd, nrow = 2, ncol = 2)

# Combine legend-less plot and legends with cowplot
plot_grid(pl, lgd_grid, rel_widths = c(3, 1))
```

```{r, fig.width=7, fig.height=8, fig.alt='The same heatmap as the previous plot. The legends are now horizontal and split into two groups, the Pearson r legend above and the annotation legends below the heatmap.'}

# Code for the same layout as above
# # Arrange in a grid (patchwork)
# lgd_grid <- wrap_plots(lgd)
# 
# # Combine with patchwork
# wrap_plots(pl, lgd_grid, widths = c(3, 1))

# Try a different layout with patchwork. Turn the legends horizontal
lgd <- get_legend(p + theme(legend.direction = "horizontal",
                            legend.title = element_text(vjust = 0.8)))[["grobs"]][1:4]

# Append the plot to the legend list to give to patchwork::wrap_plots
wrap_plots(append(list(pl), lgd),
           design = "
           ##BBB##
           AAAAAAA
           AAAAAAA
           AAAAAAA
           AAAAAAA
           CCDDEEE
           ")

```

