---
title: "Dealing with legends"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, results='hide', message=FALSE, warning=FALSE}
library(ggcorrheatmap)
library(ggplot2)
library(patchwork)  # Arrange plots and legends
library(cowplot)    # For extracting and arranging legends
library(tibble)     # For colnames with numbers
```

# Legend order

Legends have a set default order:

1. Main heatmap colours.
2. Main heatmap sizes.
3. Row annotations in order.
4. Column annotations in order.

```{r, fig.alt='A heatmap of the first 10 rows of mtcars (scaled), with coloured diamonds instead of cells. The heatmap colours range from dark blue to light blue. There is one row annotation (using the viridis D option) and one column annotation (using the viridis A option), named 3 and 4, respectively. Four legends are on the right side of the plot for the main heatmap colours, main heatmap sizes, row annotations, and column annotations. The legends are named 1, 2, 3, and 4.'}
gghm(scale(mtcars)[1:10, ], mode = 23,
     colr_name = "1", size_name = "2",
     annot_rows_df = tibble(.names = rownames(mtcars)[20:30],
                            "3" = 1:11),
     annot_cols_df = tibble(.names = colnames(mtcars),
                            "4" = 11:1)) +
  # Make the legends horizontal to fit better
  theme(legend.direction = "horizontal")
```

If the layout is mixed, the legends of the first triangle appear before those of the second triangle.

```{r, fig.alt='mtcars correlation heatmap with cells in the bottom left triangle and squares with sizes varying with the value in the top right. The deafult ggplot2 colour scale (dark blue to light blue) is used for both triangles. Three legends are on the right, the first one a colour scale, the second one a size scale, and the third one another colour scale.'}
gghm(cor(mtcars), layout = c("tr", "bl"), mode = c("15", "hm")) +
  theme(legend.direction = "horizontal")
```

The order can be changed by using the `legend_order` argument. This argument takes a numeric vector, the first element specifying the order for the first legend (of the default order), the second element for the second legend and so on. If a value is `NA`, that legend is hidden.

```{r, fig.width=8, fig.height=10, fig.alt='Two mtcars correlation heatmaps with circles instead of cells. Both have two row and two column annotations each. The first heatmap has seven legends on the right (colour, size, colour, row annotation colours 1-2, column annotation colours 1-2), named 1 to 7. The second heatmap is the same as the first but with the legends ordered 4, 2, 7, 1, 5. The 3 and 6 legends are missing.'}
set.seed(123)
annot_rows <- tibble(.names = colnames(mtcars),
                     "4" = 1:11, "5" = 11:1)
annot_cols <- tibble(.names = colnames(mtcars),
                     "6" = sample(letters[1:3], 11, TRUE),
                     "7" = sample(LETTERS[1:2], 11, TRUE))

plt1 <- gghm(cor(mtcars), layout = c("tl", "br"), mode = c(19, 21),
             colr_name = c("1", "3"), size_name = c("2"),
             annot_rows_df = annot_rows,
             annot_cols_df = annot_cols)

plt2 <- gghm(cor(mtcars), layout = c("tl", "br"), mode = c(19, 21),
             colr_name = c("1", "3"), size_name = c("2"),
             annot_rows_df = annot_rows,
             annot_cols_df = annot_cols,
             legend_order = c(4, 2, NA, 1, 5, NA, 3))

plt1 / plt2 &
  theme(legend.direction = "horizontal")
```

A single NA value hides all legends.

```{r, fig.alt='The mtcars correlation heatmap with circles as above, but with no legends.'}
gghm(cor(mtcars), layout = c("tl", "br"), mode = c(19, 21),
     colr_name = c("1", "3"), size_name = c("2"),
     annot_rows_df = annot_rows,
     annot_cols_df = annot_cols,
     legend_order = NA)
```

Scales can be specified as `ggplot2` scale objects. In this case the user has complete control over the scale and it is unaffected by arguments like `legend_order` or `colr_name`.

```{r, fig.alt='mtcars correlation heatmap with circles like above but with some colours altered. The top left triangle uses the dark to light blue colour scale, while the bottom right triangle uses the Viridis mako scale (black to teal or emerald-like green). The second row annotation uses the Brewer Purples scale (white to purple). The second column annotation has red, white, and blue. Only the bottom right triangle and second row annotation have legends to the right.'}
gghm(cor(mtcars), layout = c("tl", "br"), mode = c(19, 21),
     colr_name = c("1", "3"), size_name = "2",
     colr_scale = list(
       # Keep default scale for first triangle
       NULL,
       # Pass a scale object for second triangle
       scale_fill_viridis_c(name = "scale object 1", option = "G",
                            guide = guide_colourbar(order = 1))
     ),
     annot_rows_df = annot_rows,
     annot_cols_df = annot_cols,
     annot_rows_fill = list(
       # Scale object for second row annotation
       "5" = scale_fill_distiller(name = "scale object 2",
                                  palette = "Purples")
     ),
     annot_cols_fill = list(
       # Scale object for first column annotation
       "6" = scale_fill_brewer(name = "scale object 3",
                               palette = "RdBu",
                               # Hide the legend
                               guide = "none")
     ),
     # (Try to) Remove all legends
     legend_order = NA) +
  theme(legend.direction = "horizontal")
```

# Legends in correlation heatmaps

`ggcorrhm()` tries to hide some legends by default when using mixed layouts to reduce clutter.

For example, when both triangles use the same colour scale, one of the scales is hidden. This only happens when the default scale is used (customised with `high`, `mid`, etc is ok) or when `colr_scale` is a character. If `colr_scale` contains `ggplot2` scale objects both legends will be shown (unless specified not to in the scale object).

Default size legends are also hidden as they use an absolute value transformation and show different information from the colour scale.

```{r, fig.alt='mtcars correlation heatmap with cells in the top left and diamonds in the bottom right. The heatmap uses the Brewer RdYlBu scale that ranges from blue to yellow to red. There is only one legend on the right, showing the colour scale.'}
# If both triangles use the same colour scale
# (when colr_scale is NULL or a one character value)
ggcorrhm(mtcars, layout = c("tl", "br"),
         mode = c("hm", "18"), colr_scale = "RdYlBu")
```

The legends can be included again by using the `legend_order` argument or using scale objects for the scales.

```{r, fig.alt='mtcars correlation heatmap with cells and diamonds and the RdYlBu scale like above. There are three legends one the right, the first two for the colour scale and the last one for the sizes (showing the absolute values).'}
ggcorrhm(mtcars, layout = c("tl", "br"),
         mode = c("hm", "18"), colr_scale = "RdYlBu",
         # Elements beyond the number of legends are ignored
         # On a related note, if legend_order is shorter than
         # the number of legends, legends not in the vector are hidden
         legend_order = 1:10)
```

# Arranging legends

When there are many legends, some of them may be cut off and it can be tricky to make them fit. One way is to change the positions, sizes or orientations of legends as shown in some of the plots above. A more advanced way is to use packages like `cowplot` and `patchwork` to stitch together a legendless version of the plot with the legends, which also makes it possible to arrange the legends in grids.

```{r, fig.alt='mtcars correlation heatmap with clustered rows and columns, as well as two row annotations and one column annotation. Four legends can be seen to the right, but they do not all fit and the first and last one cannot be seen fully.'}

# Plot with legends
p <- ggcorrhm(mtcars, cluster_rows = TRUE, cluster_cols = TRUE,
              annot_rows_df = data.frame(.names = colnames(mtcars),
                                         a = 1:11, b = 11:1),
              annot_cols_df = data.frame(.names = colnames(mtcars),
                                         c = 1:11))

# Plot without legends
pl <- ggcorrhm(mtcars, cluster_rows = TRUE, cluster_cols = TRUE,
               annot_rows_df = data.frame(.names = colnames(mtcars),
                                          a = 1:11, b = 11:1),
               annot_cols_df = data.frame(.names = colnames(mtcars),
                                          c = 1:11),
               legend_order = NA)

# Show plot with default legends
p
```

```{r, fig.alt='The same heatmap with annotations as the previous figure, but the legends are now all visible, arranged in a 2x2 grid to the right of the plot.'}
# Extract the four legends
lgd <- get_legend(p)[["grobs"]][1:4]

# Arrange in a grid (cowplot)
lgd_grid <- plot_grid(plotlist = lgd, nrow = 2, ncol = 2)

# Combine legend-less plot and legends with cowplot
plot_grid(pl, lgd_grid, rel_widths = c(3, 1))
```

```{r, fig.width=7, fig.height=8, fig.alt='The same heatmap as the previous plot. The legends are now horizontal and split into two groups, the Pearson r legend above and the annotation legends below the heatmap.'}

# Code for the same layout as above
# # Arrange in a grid (patchwork)
# lgd_grid <- wrap_plots(lgd)
# 
# # Combine with patchwork
# wrap_plots(pl, lgd_grid, widths = c(3, 1))

# Try a different layout with patchwork. Turn the legends horizontal
lgd <- get_legend(p + theme(legend.direction = "horizontal",
                            legend.title = element_text(vjust = 0.8)))[["grobs"]][1:4]

# Append the plot to the legend list to give to patchwork::wrap_plots
wrap_plots(append(list(pl), lgd),
           design = "
           ##BBB##
           AAAAAAA
           AAAAAAA
           AAAAAAA
           AAAAAAA
           CCDDEEE
           ")

```

