---
title: "Mixed layouts"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, results='hide', message=FALSE, warning=FALSE}
library(ggcorrheatmap)
library(patchwork)
```

# Basics

If the matrix is symmetric it is possible to plot mixed layouts by giving two triangles (opposite each other) to the `layout` argument. Two modes must be provided, the default modes being `heatmap` for the first element and `text` for the second. The first element of `layout` gets the diagonal (if it is drawn).

```{r, fig.width=10, fig.height=5, fig.alt='Two mtcars correlation heatmaps. The first one is composed of a normal heatmap with filled cells for the top left triangle and cells with coloured text showing the correlation values in the bottom right triangle (the colour scaling with the correlation values). Names are written on the diagonal. The second heatmap has circles in the top left and diamonds in the bottom right, all filled with the same colours in the first heatmap and their sizes scaling with the absolute values of the correlations.'}
plt1 <- ggcorrhm(mtcars, layout = c("tl", "br"))
plt2 <- ggcorrhm(mtcars, layout = c("tl", "br"), mode = c("21", "23"))

plt1 + plt2
```

To add a grid for shape modes use `ggplot2::theme` and the `panel.grid*` arguments. If any part of the heatmap uses the 'text' or 'none' modes, the grid will be visible through the cells. The `cell_bg_col` and `cell_bg_alpha` arguments control the cell background colour and alpha parameter. By default, the cell colour is set to white and the alpha to 0.

```{r, fig.width=10, fig.height=5, fig.alt='Two mtcars correlation heatmaps with circles in the top left and cells with correlation labels (colour scaling with correlation value) in the bottom right. There is a major grid added to the plot, so that lines intersect at the middle points of each cell. In the first plot the grid is visible through the cells in the bottom right, making it hard to read the text. In the second plot the grid is not visible through cells.'}
# Grid + default cell background
plt1 <- ggcorrhm(mtcars, layout = c("tl", "br"), mode = c("21", "text")) +
  labs(title = "cell_bg_alpha = 0") +
  # Add grid, make darker for demonstration
  theme(panel.grid.major = element_line(colour = "grey"))

plt2 <- ggcorrhm(mtcars, layout = c("tl", "br"), mode = c("21", "text"),
                 cell_bg_alpha = 1) +
  labs(title = "cell_bg_alpha = 1") +
  theme(panel.grid.major = element_line(colour = "grey"))

plt1 + plt2
```


Since mixed layouts consist of two triangular layout heatmaps, clustering can only be applied if both rows and columns are clustered (like for the simple triangular layouts).

```{r, fig.alt='mtcars correlation heatmap with a normal heatmap in the top right and coloured text in the bottom left. Rows and columns have been clustered and dendrograms are shown on the right and below.'}
ggcorrhm(mtcars, layout = c("tr", "bl"), cluster_rows = TRUE)
```

If `return_data` is `TRUE` there will be an extra column called `layout`, showing which triangle each cell belongs to.

```{r}
head(ggcorrhm(mtcars, layout = c("tl", "br"), return_data = TRUE)$plot_data)
```

# Parameter behaviour

When the layout is mixed, a few arguments can take vectors or lists of length two. If a vector, each element is applied to the whole corresponding triangle in `layout`. If a list is provided, the elements can be whatever values that would be applicable for non-mixed layouts, i.e. a single value to apply to the whole triangle, or a vector with one value for each cell. 

The arguments that work like this are `border_col`, `border_lwd`, `border_lty`, `cell_labels`, `cell_label_col`, `cell_label_size`, `cell_label_digits`, `cell_bg_col`, `cell_bg_alpha`, and (for `ggcorrhm`) `p_values` and `cell_label_p`.

```{r, fig.alt='mtcars correlation heatmap with filled cells with thick, grey, dotted borders on the top right. The bottom left, including the diagonal with names, has cells with coloured text and thick coloured borders with a rainbow scale.'}
colr <- c("#70369D", "#59369D", "#4A44AB", "#486ED8", "#5B9892", "#79C314", "#C6DB28", "#FBDD2B", "#FEB30A", "#F56B08", "#E81416")

ggcorrhm(mtcars, layout = c("bl", "tr"), mode = c("none", "hm"),
         # One value for the whole heatmap (border linewidth)
         border_lwd = 1.5,
         # Length two vector, each value being used for the whole corresponding triangle
         border_lty = c(1, 3),
         # A list with values for each triangle
         border_col = list(
           # The first triangle is 66 values (if the diagonal is drawn), second 55
           # If not known, can return the plotting data to see how many cells are in each triangle
           # Since the matrix must be symmetric the numbers can also be calculated
           # Total: ncol*ncol, triangle without diagonal: (total - ncol) / 2, with diagonal: without + ncol
           # First triangle, provide all values, moving down each column from the left
           rev(unlist(sapply(seq_along(colr), function(i) rep(colr[i], i)))),
           # Second triangle, recycle one value
           "white"
         ),
         # Also draw cell labels on the first triangle and set the same colours as the cells
         cell_labels = c(TRUE, FALSE),
         cell_label_col = list(
           # Provide the same values as for the cell border colours, but minus one as the diagonal
           # names are written instead of cell labels and are controlled differently
           rev(unlist(sapply(seq_along(colr), function(i) rep(colr[i], i - 1)))),
           1 # Anything as there are no labels
         ),
         names_diag_param = list(colour = rev(colr)))
```

As mentioned above, p-values can be added to only one half of the heatmap if desired. P-values in correlation matrices and mixed layouts are explored more in the [correlation matrix article](correlation.html#p-values).

```{r, fig.width=10, fig.height=5, fig.alt='Two mtcars correlation heatmaps, filled cells on the top left and coloured text (showing correlation values) on the bottom right. The first plot has asterisks showing p-vaule thresholds in both triangles (added to the correlation values in the bottom right). The second plot retains the labels in the bottom right triangle but not the top left.'}
plt1 <- ggcorrhm(mtcars, layout = c("tl", "br"), p_values = TRUE,
                 cell_label_size = c(4, 2))
plt2 <- ggcorrhm(mtcars, layout = c("tl", "br"), p_values = c(FALSE, TRUE),
                 cell_label_size = 2)

plt1 + plt2
```

# Some extra plots

Using mixed layouts with mode "none" it is possible to "hack" annotations and dendrograms onto the empty sides of triangular layouts.

```{r, fig.alt='A mtcars correlation heatmap showing only the top left triangle. On the right and below are row and column annotations, leaving a large triangular gap in the bottom right.'}
annot <- data.frame(.names = colnames(mtcars), a = 1:11)
ggcorrhm(mtcars, layout = c("tl", "br"), mode = c("hm", "none"),
         annot_rows_df = annot, annot_cols_df = annot,
         cluster_rows = TRUE, cluster_cols = TRUE,
         # Hide cell borders only in the 'none' triangle
         border_lwd = c(0.5, 0))
```

Finally, the default full layout reverses the y axis levels so that the rows are in the same order as in the input and the diagonal runs from top left to bottom right. If the opposite layout is desired it can be made using a mixed layout of a top left and a bottom right triangle, like the plot below.

```{r, fig.width=10, fig.height=5, fig.alt='Two mtcars correlation heatmaps, both showing the whole heatmap with names along the diagonal. The first one has the diagonal running from top left to bottom right, while the second one has the diagonal between the bottom left and top right (the row order is reversed).'}
plt1 <- ggcorrhm(mtcars, layout = "f")
plt2 <- ggcorrhm(mtcars, layout = c("tl", "br"), mode = c("hm", "hm"))

plt1 + plt2
```

